--- 
title: "An Introduction to Bayesian Data Analysis with Stan"
author: "Daniele Bottigliengo"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Notes and exercises from the course on Bayesian Data Analysis with Stan taken during the 1st International Summer School at the Department of Brain and Behavioral Sciences Medical Statistics and Genomic Unit, University of Pavia, Italy."
---

# Information on the course {-}

This book contains notes and exercises from the course on [__Bayesian 
Data Analysis with Stan__](http://ssnc.unipv.it/msmg/images/pagine/News/Bayesian_summer_school.pdf) taken during the 1st International Summer 
School at the Department of Brain and Behavioral Sciences Medical
Statistics and Genomic Unit, University of Pavia, Italy.

The course was taught by Jonah Gabry, from [Columbia University](https://www.columbia.edu/), and Leonardo Egidi, from 
[University of Trieste](https://www.units.it/en).

## Settings {-}

Here, there are the libraries loaded during the course, w/ the relative options,
plus some packages and options useful to write code more understandable by
humans obtaining nicer output.

```{r setup, message=FALSE, warning=FALSE}
# Packages for the analyses
library(rstan)                                     # R interface to Stan

# Package(s) for data management 
library(tidyverse)            # Imports the principal tidyverse packages

# Document output options
knitr::opts_chunk$set(
  comment     = '#>',
	echo        = TRUE,                              # Render all the code
	message     = FALSE,                          # Do net render messages
	warning     = FALSE,                          # Do not render warnings
	fig.height  = 4.4,         # height to allow two figures in a PDF page
	cache.extra = knitr::rand_seed # cache seeds to assure reproducibility
)
```

The following code create the packages.bib files which is the BibTeX lists of
all the packages references we have loaded.

```{r}
# Automatically create a bib database for the loaded packages
knitr::write_bib(
  c(
    .packages(), 'bookdown', 'knitr', 'rmarkdown', 'knitr',
    'tidyverse', 'rstan'
  ),
  file = 'packages.bib'
)
```
